Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs

### 2021-02-17 submission/manuscript.Rmd
Create title, abstract, and importance sections

	if(is.numeric(x)){
		if(abs(x - round(x) < .Machine$double.eps)){
			# treat as integer
			formatted <- format(x, digits=0, big.mark = ",")

@@

# Amplicon sequence variants should not replace operational taxonomic units in marker-gene data analysis

@@

**Running title:** ASVs vs. OTUs

@@

## Abstract (250 words)

## Importance (150 words)

```{r}
metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character()))

@@

single_nasvs <- asvs_per_species %>% filter(species == single_copy$species) %>% pull(n_asvs)
ecoli_nasvs <- asvs_per_species %>% filter(species == "Escherichia coli") %>% pull(n_asvs)
```

```{r}
thresholds <- read_tsv(here("data/processed/thresholds_for_single_otu.tsv"),
											 col_types=cols(region = col_character(),
											 							 .default = col_double())) %>%
								filter(n_genomes > 100)

@@

three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19") %>% tail(n=1) %>% pull(n_rrns)
three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4") %>% tail(n=1) %>% pull(n_rrns)
three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34") %>% tail(n=1) %>% pull(n_rrns)
three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45") %>% tail(n=1) %>% pull(n_rrns)
```

```{r}
lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
											 col_types = cols(region = col_character(),
											 								 .default = col_double()))

@@

>>>>

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps){
			# treat as integer
			formatted <- format(x, digits=0, big.mark = ",")

@@

# Amplicon sequence variants artificially split bacterial genomes into separate units of inference

@@

**Running title:** ASVs artificially split bacterial genomes

@@

```{r}
## Paragraph 1
metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character()))

@@

single_nasvs <- asvs_per_species %>% filter(species == single_copy$species) %>% pull(n_asvs)
ecoli_nasvs <- asvs_per_species %>% filter(species == "Escherichia coli") %>% pull(n_asvs)

## Paragraph 2
thresholds <- read_tsv(here("data/processed/thresholds_for_single_otu.tsv"),
											 col_types=cols(region = col_character(),
											 							 .default = col_double())) %>%
								filter(n_genomes > 100)

@@

three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19") %>% tail(n=1) %>% pull(n_rrns)
three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4") %>% tail(n=1) %>% pull(n_rrns)
three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34") %>% tail(n=1) %>% pull(n_rrns)
three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45") %>% tail(n=1) %>% pull(n_rrns)

## Paragraph 3
lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
											 col_types = cols(region = col_character(),
											 								 .default = col_double()))

@@

## Abstract

Amplicon sequencing variants (ASVs) have been proposed as an alternative to operational taxonomic units (OTUs) for analyzing microbiomes. ASVs have grown in popularity, in part, because of a desire to reflect a more refined level of taxonomy because they do not cluster sequences based on a distance-based threshold. However, ASVs and the use of overly narrow thresholds to identify OTUs increases the risk of splitting a single genome into separate clusters. I addressed this problem by analyzing the intragenomic variation of 16S rRNA genes from the bacterial genomes represented in the rrn copy number database, which contained `r nrow(metadata)` genomes from `r n_distinct(metadata$species)` species. The analysis confirmed the severity of this risk. As the number of copies of the 16S rRNA gene increased in a genome, the number of ASVs also increased. For full length 16S rRNA genes, there was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene. Among genomes with 7 copies of the 16S rRNA, such as *E. coli*, it was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster the ASVs from the same genome into a single OTU. This research highlights the risk of splitting a single bacterial genome into separate clusters when ASVs are used to analyze 16S rRNA gene sequence data. Although there is also a risk of clustering different species into the same OTU, the effects of those risks on biological inferences are less than those from artificially splitting a genome into separate ASVs and OTUs.

## Importance

16S rRNA gene sequencing has propelled significant interest into host-associated and environmental microbiomes. There is a tension between trying to classify 16S rRNA gene sequences to increasingly lower taxonomic levels and the reality that those levels were defined using more sequence and physiological information than is available from a fragment of the 16S rRNA gene. Furthermore, naming of bacterial taxa reflects the biases of those who name them. One motivation for the recent push to adopt ASVs in place of OTUs in microbiome analyses is to allow researchers to perform their analyes at the finest possible level that reflects species-level taxonomy. The current research is significant because it quantifies the risk of artificially splitting bacterial genomes into separate clusters. Far from providing a better represenation of microbiomes, the ASV approach could lead to conflicting inferences about the ecology of different clusters from the same genome.

### 2021-02-17 submission/manuscript.tex
Create title, abstract, and importance sections

\hypertarget{amplicon-sequence-variants-should-not-replace-operational-taxonomic-units-in-marker-gene-data-analysis}{%
\section{Amplicon sequence variants should not replace operational
taxonomic units in marker-gene data
analysis}\label{amplicon-sequence-variants-should-not-replace-operational-taxonomic-units-in-marker-gene-data-analysis}}

@@

\textbf{Running title:} ASVs vs.~OTUs

@@

\hypertarget{abstract-250-words}{%
\subsection{Abstract (250 words)}\label{abstract-250-words}}

\hypertarget{importance-150-words}{%
\subsection{Importance (150 words)}\label{importance-150-words}}

@@

To investigate the variation in the number of copies of the 16S rRNA
gene per genome as well as the intragenomic variation among copies of
the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
copy number database (rrnDB)(14). Among the \textbf{4,774} species
represented in the rrnDB there were \textbf{15,614} genomes. The median
number of rrn operson per species ranged between \textbf{1} (e.g.,
\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
(\textbf{\emph{Metabacillus litoralis}}) copies of the rrn operon. As
the number of copies of the operon in a genome increased, the number of
variants of the 16S rRNA gene in each genome also increased. On average,
there were \textbf{1} variants per copy of the full length 16S rRNA gene
and an average of \textbf{0.26}, \textbf{0.33}, and \textbf{0.27}
variants when considering the V4, V3-V4, and V4-V5 regions of the gene,
respectively. Although a species tended to have a consistent number of
16S rRNA gene copies per genome, the number of total variants increased
with the number of genomes that were sampled (\textbf{Figure 1}). For
example, \textbf{\emph{Mycobacterium tuberculosis}} generally only had
\textbf{1} copy of the gene per genome, but across the \textbf{180}
genomes that have been sequenced there were \textbf{11} versions of the
gene. Similarly, a \emph{E. coli} genome typically had \textbf{7} copies
of the 16S rRNA gene with between \textbf{6} and \textbf{10} distinct
full length sequences per genome. Across the \textbf{958} \emph{E. coli}
genomes that have been sequenced, there were \textbf{1,013} different
variants of the gene. These observations highlight the risk of selecting
a threshold for defining units of inference that is too narrow because
it is possible to split a single genome into multiple units.

>>>>

\hypertarget{amplicon-sequence-variants-artificially-split-bacterial-genomes-into-separate-units-of-inference}{%
\section{Amplicon sequence variants artificially split bacterial genomes
into separate units of
inference}\label{amplicon-sequence-variants-artificially-split-bacterial-genomes-into-separate-units-of-inference}}

@@

\textbf{Running title:} ASVs artificially split bacterial genomes

@@

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

Amplicon sequencing variants (ASVs) have been proposed as an alternative
to operational taxonomic units (OTUs) for analyzing microbiomes. ASVs
have grown in popularity, in part, because of a desire to reflect a more
refined level of taxonomy because they do not cluster sequences based on
a distance-based threshold. However, ASVs and the use of overly narrow
thresholds to identify OTUs increases the risk of splitting a single
genome into separate clusters. I addressed this problem by analyzing the
intragenomic variation of 16S rRNA genes from the bacterial genomes
represented in the rrn copy number database, which contained
\textbf{15,614} genomes from \textbf{4,774} species. The analysis
confirmed the severity of this risk. As the number of copies of the 16S
rRNA gene increased in a genome, the number of ASVs also increased. For
full length 16S rRNA genes, there was an average of \textbf{0.60} ASVs
per copy of the 16S rRNA gene. Among genomes with 7 copies of the 16S
rRNA, such as \emph{E. coli}, it was necessary to use a distance
threshold of \textbf{5.5}\% to cluster the ASVs from the same genome
into a single OTU. This research highlights the risk of splitting a
single bacterial genome into separate clusters when ASVs are used to
analyze 16S rRNA gene sequence data. Although there is also a risk of
clustering different species into the same OTU, the effects of those
risks on biological inferences are less than those from artificially
splitting a genome into separate ASVs and OTUs.

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

16S rRNA gene sequencing has propelled significant interest into
host-associated and environmental microbiomes. There is a tension
between trying to classify 16S rRNA gene sequences to increasingly lower
taxonomic levels and the reality that those levels were defined using
more sequence and physiological information than is available from a
fragment of the 16S rRNA gene. Furthermore, naming of bacterial taxa
reflects the biases of those who name them. One motivation for the
recent push to adopt ASVs in place of OTUs in microbiome analyses is to
allow researchers to perform their analyes at the finest possible level
that reflects species-level taxonomy. The current research is
significant because it quantifies the risk of artificially splitting
bacterial genomes into separate clusters. Far from providing a better
represenation of microbiomes, the ASV approach could lead to conflicting
inferences about the ecology of different clusters from the same genome.

@@

To investigate the variation in the number of copies of the 16S rRNA
gene per genome as well as the intragenomic variation among copies of
the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
copy number database (rrnDB)(14). Among the \textbf{4,774} species
represented in the rrnDB there were \textbf{15,614} genomes. The median
number of rrn operson per species ranged between \textbf{1} (e.g.,
\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
(\textbf{\emph{Metabacillus litoralis}}) copies of the rrn operon. As
the number of copies of the operon in a genome increased, the number of
variants of the 16S rRNA gene in each genome also increased. On average,
there were \textbf{0.60} variants per copy of the full length 16S rRNA
gene and an average of \textbf{0.26}, \textbf{0.33}, and \textbf{0.27}
variants when considering the V4, V3-V4, and V4-V5 regions of the gene,
respectively. Although a species tended to have a consistent number of
16S rRNA gene copies per genome, the number of total variants increased
with the number of genomes that were sampled (\textbf{Figure 1}). For
example, \textbf{\emph{Mycobacterium tuberculosis}} generally only had
\textbf{1} copy of the gene per genome, but across the \textbf{180}
genomes that have been sequenced there were \textbf{11} versions of the
gene. Similarly, a \emph{E. coli} genome typically had \textbf{7} copies
of the 16S rRNA gene with between \textbf{6} and \textbf{10} distinct
full length sequences per genome. Across the \textbf{958} \emph{E. coli}
genomes that have been sequenced, there were \textbf{1,013} different
variants of the gene. These observations highlight the risk of selecting
a threshold for defining units of inference that is too narrow because
it is possible to split a single genome into multiple units.
